Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-22 paper-filmat.tex
typo

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.
% XXX: mention Olivia?
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
% XXX contingent: To some extent, what the commonly agreed-upon rules of
% mathematics are is dependent on contingent historical factors. It is
% conceivable, for instance, that the intuitionists ...
\end{enumerate}
This note touches on these topics and invites further research.

@@

Formally, a topos is a certain kind of \emph{category}, containing objects and
morphisms between those objects. The formal definition, recorded here only for
reference, requires some amount of category theory, but, as will be outlined in
the following sections, exploring the mathematical universe of a given topos
does not.

@@

\begin{defn}The \emph{standard topos}~$\Set$ is the category which contains all
sets as its objects and all maps between sets as morphisms.\end{defn}

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
translating conjunction reads
\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
The remaining translation rules are more involved; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes.

@@

The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
support intuitionistic reasoning, not the full power of the ordinary
\emph{classical reasoning}. That is, within most toposes, the law of excluded
middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
($\neg\neg\varphi \Rightarrow \varphi$) are not available. It is intuitionistic
logic and not classical logic which is the common denominator of all toposes.

While it may appear that these two laws pervade any mathematical theory, in
fact a substantial amount of mathematics can be developed intuitionistically
(see for
instance~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra}
for constructive algebra,~\cite{bishop-bridges:constructive-analysis} for
constructive analysis
and~\cite{bauer:int-mathematics,bauer:video,melikhov:intuitionistic-logic} for
further references) and hence the alternate universes provided by toposes
cannot be too strange: In any topos, there are infinitely many prime numbers,
the square root of two is not rational and the powerset of the naturals is
uncountable.

@@

\begin{enumerate}
\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
is true. The trivial topos is not interesting on its own, but its existence
streamlines the theory and it can be an interesting question whether a given
topos coincides with the trivial topos.
\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
in the ordinary mathematical sense.
\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
that the Kripke--Joyal semantics of toposes generalizes the more familiar
Kripke semantics.
\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
if it has a \emph{computable witness} as detailed in
Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
choice holds (even if it does not on the meta level).
\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
failure is for geometric reasons.
\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
Section~\ref{sect:smooth}. This topos contains a mirror
image of~$A$ which is a field, even if~$A$ is not.
\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
regarded as classical mechanical systems (which are described by commutative
algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
and~$\psi$ are geometric formulas (formulas built from equality and specified
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
set-indexed disjunctions~$\bigvee$).} This topos contains the
\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
of groups contains the \emph{generic group}. Arguably it is this group
which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
The generic group has has exactly those
properties which are shared by any group whatsoever.\footnote{More precisely,
this is only true for those properties which can be formulated as geometric
sequents. For arbitrary properties~$\varphi$, the statements~``the generic
group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
not be equivalent. This situation is explored
in~\cite{blechschmidt:nullstellensatz}.}
% FUTURE: Give Set[T] its own section.
\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
intuitionistically provable. Lambek and Scott proposed that the free topos can
reconcile moderate platonism (because this topos has a certain universal
property which can be used to single it out among the plenitude of toposes),
moderate formalism (because it is constructed in a purely syntactic way) and
moderate logicism (because, as a topos, it supports an intuitionistic type
theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
\end{enumerate}

@@

Toposes are still mathematical structures, and as long as we study
toposes within the usual setup of mathematics, our toposes are all part of the
standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
topos twice.\footnote{There is a fine print to consider. Technically, if we
work within~\textsc{zf} or its intuitionistic cousin~\textsc{izf}, most toposes
of interest are proper classes, not sets. In particular~$\Set$ itself is a
proper class. Hence Figure~\ref{fig:landscape} should not be interpreted as
indicating that toposes are contained in~$\Set$ as objects, which most are not.
In this regard toposes are similar to class-sized inner models in set theory.
We believe that ``our toposes are all part of the standard topos'' is still an
apt description of the situation.}
The toposes which we can study in mathematics do not tell us
all possible stories how mathematics could unfold, only those which appear
coherent from the point of view of the standard topos, and the topos-theoretic
multiverse which we have access to is just a small part of an even larger
landscape.\footnote{This paragraph employs an overly narrow conception of
``mathematics'', for instance excluding any predicative flavors of mathematics
(Laura Crosilla's survey~\cite{crosilla:predicative} is an excellent introduction).
Toposes are impredicative in the sense that any object of a topos is required to have a
powerobject. A predicative cousin of toposes are the \emph{arithmetic
universes} introduced by Joyal which have recently been an important object of
consideration by Maietti and
Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}

@@

In such questions of computability, practical issues such as resource
constraints and hardware malfunctions are ignored; we work with the theoretical
notion of \emph{Turing machines}, a mathematical abstraction of the computers
of the real world.

@@

\bigskip
\begin{center}
\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  \bottomrule
\end{tabular}
\end{center}

@@

By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
amounts to stating that there is a Turing machine which, when given a natural
number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
not. Such a Turing machine indeed exists -- writing such a program is often a
first exercise in programming courses. Hence the statement is also true in the
\effective topos, but for the nontrivial reason that such a machine exists.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} More
formally, the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is computable.''}
The preceding examples give the impression that what is true in the
\effective topos is simply a subset of what is true in the standard topos. The
example of this subsection, the so-called \emph{formal Church--Turing thesis},
shows that the relation between the two toposes is more nuanced.

@@

In contrast, in the \effective topos, any function~$\NN \to \NN$ is computable
by a Turing machine. The external meaning of this internal statement is that
there exists a Turing machine~$M$ which, when given a description of a Turing
machine~$F$ computing a function~$f : \NN \to \NN$, outputs a description of a
Turing machine computing~$f$. It is trivial to program such a machine~$M$: the
machine~$M$ simply has to echo its input back to the caller.

@@

\bigskip
\paragraph{Hypercomputation}
Firstly, in place of ordinary Turing machines, one can employ the
\emph{infinite-time Turing machines} pioneered by Hamkins and
Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
in that they can run for ``longer than infinity''; more precisely, the
computational steps are indexed by the ordinal numbers instead of the natural
numbers. For instance, an infinite-time Turing machine can trivially decide the
twin prime conjecture, by simply walking along the natural number line and
recording any twin primes it finds. Then, on day~$\omega$, it can observe
whether it has found infinitely many twins or not.

@@

The weaker statement that for any real number~$a$ it is \notnot the case that~$a < 0$
or~$a = 0$ or~$a > 0$ does hold in~$\Sh(X)$, for this statement is an
intuitionistic tautology. Its meaning is that there exists a
dense open~$U$ such that~$U$ can be covered by opens on which~$a$ is either
everywhere negative, everywhere zero or everywhere positive. In the example
given in Figure~\ref{fig:trichotomy}(c), this open~$U$ could be taken as~$X$
with the unique zero of~$a$ removed.

@@

The idea of \emph{infinitesimal numbers} -- numbers which lie between~$-\frac{1}{n}$
and~$\frac{1}{n}$ for any natural number~$n$ -- has a long and rich history. They are
not part of today's standard setup of the reals, but they are still intriguing
as calculational tools and as a device to bring mathematical intuition and
mathematical formalism closer together.

@@

In the ``if'' direction, the transfer principle is useful for importing
knowledge about the ordinary reals in the hyperreal realm. For instance,
addition of hyperreals is commutative because addition of reals is.
By the ``only if'' direction, a theorem established for the hyperreals also
holds for the ordinary reals. In this way, the infinitesimal numbers of
nonstandard analysis can be viewed as a convenient fiction, generating a
conservative extension of the usual setup of mathematics.

@@

The first deficiency could be fixed by passing
from~$\RR[\varepsilon]/(\varepsilon^2)$ to~$\RR[\varepsilon]/(\varepsilon^3)$
(a ring whose elements are triples and whose ring operations are defined such
that~$\langle0,1,0\rangle^3 = 0$) and the second by passing
from~$\RR[\varepsilon]/(\varepsilon^2)$
to~$\RR[\varepsilon,\varepsilon']/(\varepsilon^2,\varepsilon'^2,\varepsilon\varepsilon')$.
In a sense, both of these proposed replacements are better \emph{stages} than
the basic ring~$\RR[\varepsilon]/(\varepsilon)^2$ or even~$\RR$ itself. However,
similar criticisms can be mounted against any of these better stages, and the
problem that all these substitutes are not fields persists.

@@

Within~$\Zar(\RR)$, we can construct the set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ has the following properties:
\begin{enumerate}
\item Law of cancellation: $\forall x \? \RR^\sim\_ \forall y \? \RR^\sim\_ \bigl((\forall
\varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
\Rightarrow x = y\bigr)$
\item Axiom of micro-affinity: $\forall f \? {\RR^\sim}^\Delta\_ \exists! a \? \RR^\sim\_
\forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
\end{enumerate}
The unique number~$a$ in the axiom of micro-affinity deserves to be
called~``$f'(0)$''; this is how we synthetically define the derivative in
synthetic differential geometry.

@@

Having explored the alternate universes of several toposes in
Sections~\ref{sect:effective-topos} to~\ref{sect:smooth}, we are now in a
position to explain our view of toposes as lenses in more details.

Toposes can be seen in two quite different ways:
\begin{enumerate}
\item (semantic view) A topos is an alternate mathematical universe. Any topos
contains its own stock of mathematical objects. A ``transfer theorem'' relates
properties of those objects with properties of objects of the standard topos: A
statement~$\varphi$ about the objects of a topos~$\E$ holds in~$\E$ iff the
statement~``$\E \models \varphi$'' holds in the standard topos.
\item (syntactic view) A topos is merely an index to a syntactical translation
process. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
turns a statement~$\varphi$ (about ordinary mathematical objects) into the
statement~``$\E \models \varphi$'' of the same kind (again about ordinary
mathematical objects).
\end{enumerate}

>>>>

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.
% XXX: mention Olivia?
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
% XXX contingent: To some extent, what the commonly agreed-upon rules of
% mathematics are is dependent on contingent historical factors. It is
% conceivable, for instance, that the intuitionists ...
\end{enumerate}
We invite further research on these connections.

@@

A topos is a certain kind of \emph{category}, containing objects and
morphisms between those objects. The precise definition is recorded here only for
reference. Appreciating it requires some amount of category theory, but, as will be demonstrated in
the following sections, exploring the mathematical universe of a given topos
does not.

@@

\begin{defn}The \emph{standard topos}~$\Set$ is the category which has all
sets as its objects and all maps between sets as morphisms.\end{defn}

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
translating conjunction reads
\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
The remaining translation rules are more
involved~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes.

@@

The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
support intuitionistic reasoning, not the full power of the ordinary
\emph{classical reasoning}. That is, within most toposes, the law of excluded
middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
($\neg\neg\varphi \Rightarrow \varphi$) are not available. It is intuitionistic
logic and not classical logic which is the common denominator of all toposes;
we cannot generally argue by contradiction in a topos.

While it may appear that these two laws pervade any mathematical theory, in
fact a substantial amount of mathematics can be developed intuitionistically
(see for
instance~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra}
for constructive algebra,~\cite{bishop-bridges:constructive-analysis} for
constructive analysis
and~\cite{bauer:int-mathematics,bauer:video,melikhov:intuitionistic-logic} for
accessible surveys on appreciating intuitionistic logic) and hence the alternate universes provided by toposes
cannot be too strange: In any topos, there are infinitely many prime numbers,
the square root of two is not rational, the fundamental theorem of Galois
theory holds and the powerset of the naturals is uncountable.

@@

\begin{enumerate}
\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
is true. The trivial topos is not interesting on its own, but its existence
streamlines the theory and it can be an interesting question whether a given
topos coincides with the trivial topos.
\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
in the ordinary mathematical sense.
\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
that the Kripke--Joyal semantics of toposes generalizes the more familiar
Kripke semantics.
\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
if it has a \emph{computable witness} as detailed in
Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
choice holds (even if it does not on the meta level).
\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
failure is for geometric reasons.
\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
Section~\ref{sect:smooth}. This topos contains a mirror
image of~$A$ which is a field, even if~$A$ is not.
\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
regarded as classical mechanical systems (which are described by commutative
algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
and~$\psi$ are geometric formulas (formulas built from equality and specified
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
set-indexed disjunctions~$\bigvee$).} This topos contains the
\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
of groups contains the \emph{generic group}. Arguably it is this group
which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
The generic group has has exactly those
properties which are shared by any group whatsoever.\footnote{More precisely,
this is only true for those properties which can be formulated as geometric
sequents. For arbitrary properties~$\varphi$, the statements~``the generic
group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
not be equivalent. This situation is explored
in~\cite{blechschmidt:nullstellensatz}.}
% FUTURE: Write just a little more on Set[T], that geom.impl. consequences spread.
% FUTURE: Give Set[T] its own section.
\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
intuitionistically provable. Lambek and Scott proposed that the free topos can
reconcile moderate platonism (because this topos has a certain universal
property which can be used to single it out among the plenitude of toposes),
moderate formalism (because it is constructed in a purely syntactic way) and
moderate logicism (because, as a topos, it supports an intuitionistic type
theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
\end{enumerate}

@@

Toposes are still mathematical structures, and as long as we study
toposes within the usual setup of mathematics, our toposes are all part of the
standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
topos twice.\footnote{There is a fine print to consider. Technically, if we
work within~\textsc{zf} or its intuitionistic cousin~\textsc{izf}, most toposes
of interest are proper classes, not sets. In particular~$\Set$ itself is a
proper class. Hence Figure~\ref{fig:landscape} should not be interpreted as
indicating that toposes are contained in~$\Set$ as objects, which most are not.
In this regard toposes are similar to class-sized inner models in set theory.
We believe that ``our toposes are all part of the standard topos'' is still an
apt description of the situation.}
The toposes which we can study in mathematics do not tell us
all possible stories how mathematics could unfold, only those which appear
coherent from the point of view of the standard topos, and the topos-theoretic
multiverse which we have access to is just a small part of an even larger
landscape.\footnote{This paragraph employs an overly narrow conception of
``mathematics'', focusing only on those mathematical worlds which form toposes
and for instance excluding any predicative flavors of mathematics
(Laura Crosilla's survey~\cite{crosilla:predicativity} is an excellent introduction).
Toposes are impredicative in the sense that any object of a topos is required to have a
powerobject. A predicative cousin of toposes are the \emph{arithmetic
universes} introduced by Joyal which have recently been an important object of
consideration by Maietti and
Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}

@@

In such questions of computability, practical issues such as resource
constraints or hardware malfunctions are ignored; we employ the theoretical
notion of \emph{Turing machines}, a mathematical abstraction of the computers
of the real world.

@@

\bigskip
\begin{center}
\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  % FUTURE: Countable axiom of choice
  \bottomrule
\end{tabular}
\end{center}

@@

By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
amounts to stating that there is a Turing machine which, when given a natural
number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
not. Such a Turing machine indeed exists -- writing such a program is often a
first exercise in programming courses. Hence the statement is also true in the
\effective topos, but for the nontrivial reason that primality can be
algorithmically tested.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is computable.''}
The preceding examples give the impression that what is true in the
\effective topos is solely a subset of what is true in the standard topos. The
example of this subsection, the so-called \emph{formal Church--Turing thesis},
shows that the relation between the two toposes is more nuanced.

@@

In contrast, in the \effective topos, any function~$\NN \to \NN$ is computable
by a Turing machine. The external meaning of this internal statement is that
there exists a Turing machine~$M$ which, when given a description of a Turing
machine~$F$ computing a function~$f : \NN \to \NN$, outputs a description of a
Turing machine computing~$f$. It is trivial to program such a machine~$M$: the
machine~$M$ simply has to echo its input back to the user.

@@

\bigskip
\paragraph{Hypercomputation}
Firstly, in place of ordinary Turing machines, one can employ the
\emph{infinite-time Turing machines} pioneered by Hamkins and
Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
in that they can run for ``longer than infinity''; more precisely, their
computational steps are indexed by the ordinal numbers instead of the natural
numbers. For instance, an infinite-time Turing machine can trivially decide the
twin prime conjecture, by simply walking along the natural number line and
recording any twin primes it finds. Then, on day~$\omega$, it can observe
whether it has found infinitely many twins or not.

@@

The weaker statement ``for any real number~$a$ it is \notnot the case that~$a < 0$
or~$a = 0$ or~$a > 0$'' does hold in~$\Sh(X)$, for this statement is an
intuitionistic tautology. Its meaning is that there exists a
dense open~$U$ such that~$U$ can be covered by opens on which~$a$ is either
everywhere negative, everywhere zero or everywhere positive. In the example
given in Figure~\ref{fig:trichotomy}(c), this open~$U$ could be taken as~$X$
with the unique zero of~$a$ removed.

@@

The idea of \emph{infinitesimal numbers} -- numbers which lie between~$-\frac{1}{n}$
and~$\frac{1}{n}$ for any natural number~$n$ -- has a long and rich history. They are
not part of today's standard setup of the reals, but they are still intriguing
as calculational tools and as a device to bring mathematical intuition and
mathematical formalism closer together.
% FUTURE: Give references to philosophical accounts of infinitesimal numbers.

@@

In the ``if'' direction, the transfer principle is useful for importing
knowledge about the ordinary reals into the hyperreal realm. For instance,
addition of hyperreals is commutative because addition of reals is.
By the ``only if'' direction, a theorem established for the hyperreals also
holds for the ordinary reals. In this way, the infinitesimal numbers of
nonstandard analysis can be viewed as a convenient fiction, generating a
conservative extension of the usual setup of mathematics.

@@

The first deficiency could be fixed by passing
from~$\RR[\varepsilon]/(\varepsilon^2)$ to~$\RR[\varepsilon]/(\varepsilon^3)$
(a ring whose elements are triples and whose ring operations are defined such
that~$\langle0,1,0\rangle^3 = 0$) and the second by passing
from~$\RR[\varepsilon]/(\varepsilon^2)$
to~$\RR[\varepsilon,\varepsilon']/(\varepsilon^2,\varepsilon'^2,\varepsilon\varepsilon')$.
In a sense, both of these proposed replacements are better \emph{stages} than
the basic ring~$\RR[\varepsilon]/(\varepsilon^2)$ or even~$\RR$ itself. However,
similar criticisms can be mounted against any of these better stages, and the
problem that all these substitutes are not fields persists.

@@

Within~$\Zar(\RR)$, we can construct the set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ validates the following laws:
\begin{enumerate}
\item Law of cancellation: $\forall x \? \RR^\sim\_ \forall y \? \RR^\sim\_ \bigl((\forall
\varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
\Rightarrow x = y\bigr)$
\item Axiom of micro-affinity: $\forall f \? {\RR^\sim}^\Delta\_ \exists! a \? \RR^\sim\_
\forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
\end{enumerate}
The unique number~$a$ in the axiom of micro-affinity deserves to be
called~``$f'(0)$''; this is how we synthetically define the derivative in
synthetic differential geometry.

@@

Having explored the alternate universes of several toposes in
Sections~\ref{sect:effective-topos} to~\ref{sect:smooth}, we are now in a
position to explain our view of toposes as lenses in more detail.

Toposes can be seen in two quite different ways:
\begin{enumerate}
\item (semantic view) A topos is an alternate mathematical universe. Any topos
contains its own stock of mathematical objects. A ``transfer theorem'' relates
properties of those objects with properties of objects of the standard topos: A
statement~$\varphi$ about the objects of a topos~$\E$ holds in~$\E$ iff the
statement~``$\E \models \varphi$'' holds in the standard topos.
\item (syntactic view) A topos is merely an index to a syntactical translation
procedure. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
turns a statement~$\varphi$ (about ordinary mathematical objects) into the
statement~``$\E \models \varphi$'' of the same kind (again about ordinary
mathematical objects).
\end{enumerate}
